**Title:** Analyzing and comparing long-term water quality of lakes and reservoirs in Rhode Island and the Northeastern United States using an anomaly approach

### 2020-09-25 manuscript/ecosphere_resubmit/ecosphere_response.Rmd
push before resubmit to ecosphere

---
output: 
  word_document:
    reference_docx: response_template.docx
editor_options: 
  chunk_output_type: console
---
```{r include = FALSE}
library(knitr)
library(readr)
library(dplyr)
library(tidyr)
library(lubridate)
opts_chunk$set(message = FALSE, warning = FALSE)
```
# Reviewer 1

@@

- **Response:**  Good catch!  Fortunately, it has not greatly impacted the analysis.  The table below lists the percentage of lake-years for each parameter that had at least one sample in each of May/June, July/Aug, and Sept/Oct.  While we have full seasonal coverage for the majority of lake-years we decided to follow the reviewers suggestion and filter these out.  This filtering step has been added to our `data_clean.R` script at lines 157 and 398 and is in the `filter_months()` function in `functions.R`.  This removes any lake/year/parameter that does not have a minimum of one sample in May/June, July/August, and September/October for both the Watershed Watch data and the LAGOSNE data.  The changes to the analysis are minimal and do not significantly alter any of our conclusions.  We added text in the second paragraph of the Methods section describing the selection criteria.

```{r echo = FALSE, results="asis"}
knitr::kable(param_min_samp)
```

@@

- **Response:** What we mean by this, is if sites with differing base lines (e.g. low nutrient site vs high nutrient site or cool water site vs warm water site) were included at different times, then site-specific anomalies are more appropriate for examining long-term trends because each of the sites will be re-scaled, relative to that sites long term mean.  For instance, comparing measured values from a cool water site that shows no trend which is measured mostly early in a sampling period to a warm water site without a trend measured late in a sampling period would likely show an increasing trend when in fact none existed.  Re-scaling these values relative to the mean sets both sites to a common baseline and would then not show a trend.  To further illustrate this, look at the two figures below.  The first figure shows simulated data from three sites.  All sites have no trend with site one being measured across the time period with values mostly in the middle of a range.  Site 2 has no trend with low values and is only in the first portion of the time period.  Site 3 has no trend with high values and is only in the later portion of the time period.  The second figure shows what the average yearly trend would look like if just looking at the values vs looking at the anomalies.  The measured values show a trend, when none should exist.  The anomalies correctly show no increasing trend.  We have added in these figures and text to the Discussion

@@

```{r, echo=FALSE}
examp_site_summ <- examp %>% 
  group_by(site) %>%
  mutate(lt_mean = mean(values)) %>%
  ungroup() %>%
  group_by(site, year) %>%
  summarize(measured_value = mean(values),
            anomaly = mean(values-lt_mean)) %>%
  ungroup() %>%
  pivot_longer(measured_value:anomaly, "Method")

examp_yr_summ <- examp_site_summ %>%
  group_by(year, Method) %>%
  summarize(values = mean(value)) %>%
  ungroup()

@@

- **Response:** We like the reviewers concept of a more holistic way of interpreting trends.  We attempted to do this in our initial assessment of these trends, but we obviously were not successful in doing so.  We now lay out a more explicit approach and describe how we use sign of the slope, magnitude of the slope, relative frequency of high/low years in the beginning and and end of the time frame, and the p-value together to assess for trends.  This is described in the paper in a new section in the methods, *Assessing regressions for trends*.   One of the biggest changes is to suggest values for the slope that would indicate an ecologically meaningful trend.  We describe this in the new section, but in general we are using a trophic state change from oligotrophic to eutrophic over one century to indicate ecologically meaningful change.  For temperature we look at values associated with ecological change as well as policy values (e.g. the Paris Agreement).  While these exact numbers are not universal, we use them as a starting point for assessing the trends.  How we interpreted the trends relative to the additional specific issues raised by the reviewer are detailed below and have changed somewhat from our initial submission, altough the broad patterns still hold.

@@

- **Response:**  The results for TN in Rhode Island have changed somewhat since we added the additional filtering steps suggested by Reviewer 1 (*1-1* and *1-2*) and made changes to use the median instead of the mean in our response to comment *1-6*.  These changes have weakened the positive trend as the slope is smaller and the p-value larger.  Additionally three years, 1995, 1996, and 1997 now have insufficient data to be included in the figures and a third, 1998, has only three sites.  Using the more explicit definition of a trend we now conclude that TN showed neither a statistically positive trend or an ecologically important trend over the time frame of a century as the slope is less than the amount needed to cause an trophic state change.  This is reflected in the manuscript.

@@

- **Response:** We applied the same within and across year filters to the LAGOSNE, used the median instead of the mean and these results have changed as well.  Given that and our method for interpreting the trends leads to the following:  LAGOSNE is showing almost no statistical support for a negative change in nitrogen, with more below average years recently.  The slope is much less than the 3.5 we identified.  Thus we conclude no trend in TN at the scale of the Region.  This is indicated in the manuscript.

@@

- **Response:** Again, with the changes to the filtering the trend lines have slightly changed. TP is showing no trend with no statistical support for a trend, a very small slope and although above average years are evently distributed.  We have indicated as such in the text.

@@

- **Response:** We were a little unclear about this comment.  Is it referring to coherence over time within a single variable or coherence between the variables?  We agree that for individual variables some years do show greater variability than others.  These increases in the yearly standard deviations are largely the result of a few unusual measurements in those years for those variables.  Essentially these are outliers, but ones that have no obvious reason to remove from the analysis so they are kept in.  Further, there is no pattern to these outliers as they are from different ponds, aren't seen across variables in the same years, and in the case of the LAGOSNE data, they come from different states.  There is only one exception to this and that is the high variability in 2011 for both TN and TP.  That is driven by one sampling event where TN and TP were high for both.  Given the lack of a pattern, there is little additional inference we can draw from this, other than the analysis might be susceptible to outliers.  Given that our original analysis is based on the mean and std. dev. and that those statistics are sensitive to outliers we have decided, thanks to this comment, to change to median and the inter-quartile range, both of which are considerably more robust statistics in the presence of outliers.  With this change the swings in variance are no longer seen as would be expected with statistics that are robust to outliers.  Good catch, by the reviewer.

@@

- **Response:** We agree with the reviewer that lake scale trends and the drivers behind those trends are both very interesting questions and something our approach could contribute to; however, we feel that it is beyond the scope of this particular paper.  Given the changes we have made to the title and the stated goals of the paper we feel that adding this type of analysis would greatly expand the paper and take away from our primary purposes, looking at long term trends in Rhode Island and the Northeast/Midwest.  We do feel this is a very good suggestion for next steps and my co-authors and I are in active discussion on how to move forward with such an analysis.   In the Discussion at the the end of *Trends* we briefly mention this application and that the anomaly approach could also be used at the individual lake level to assess for trends.

>>>>

---
output: 
  word_document:
    reference_docx: response_template.docx
editor_options: 
  chunk_output_type: console
---
```{r include = FALSE}
library(knitr)
library(readr)
library(dplyr)
library(tidyr)
library(lubridate)
opts_chunk$set(message = FALSE, warning = FALSE)
```

We would like to extend our thanks to the reviewers for two high quality reviews.  We found the comments to be thoughtful and in all cases the final product, we feel, is much better that what we originally submitted.  Below you will find our itemized response to each comment made by the reviewers.

# Reviewer 1

@@

- **Response:**  Good catch!  Fortunately, it has not greatly impacted the analysis as we have full seasonal coverage for the majority of lake-years.  That being said, we still decided to follow the reviewer's suggestion and filter these out.  This filtering step has been added to our `data_clean.R` script at lines 157 and 398 and is in the `filter_months()` function in `functions.R`.  This removes any lake/year/parameter that does not have a minimum of one sample in May/June, July/August, and September/October for both the Watershed Watch data and the LAGOSNE data.  The changes to the analysis are minimal and do not significantly alter any of our conclusions.  We added text in the second paragraph of the Methods section describing the selection criteria.

@@

- **Response:** What we mean by this, is if sites with differing base lines (e.g. low nutrient site vs high nutrient site or cool water site vs warm water site) were included at different times, then site-specific anomalies are more appropriate for examining long-term trends because each of the sites will be re-scaled, relative to that sites long term median.  For instance, comparing measured values from a cool water site that shows no trend which is measured mostly early in a sampling period to a warm water site without a trend measured late in a sampling period would likely show an increasing trend when in fact none existed.  Re-scaling these values relative to the median sets both sites to a common baseline and would then not show a trend.  To further illustrate this, look at the two figures below.  The first figure shows simulated data from three sites.  All sites have no trend with site one being measured across the time period with values mostly in the middle of a range.  Site 2 has no trend with low values and is only in the first portion of the time period.  Site 3 has no trend with high values and is only in the later portion of the time period.  The second figure shows what the average yearly trend would look like if just looking at the values vs looking at the anomalies.  The measured values show a trend, when none should exist.  The anomalies correctly show no increasing trend.  We have added in similar figures and text to the Discussion

@@

```{r, echo=FALSE}
examp_site_summ <- examp %>% 
  group_by(site) %>%
  mutate(lt_median = median(values)) %>%
  ungroup() %>%
  group_by(site, year) %>%
  summarize(measured_value = median(values),
            anomaly = median(values-lt_median)) %>%
  ungroup() %>%
  pivot_longer(measured_value:anomaly, "Method")

examp_yr_summ <- examp_site_summ %>%
  group_by(year, Method) %>%
  summarize(values = median(value)) %>%
  ungroup()

@@

- **Response:** We like the reviewers concept of a more holistic way of interpreting trends.  We attempted to do this in our initial assessment of these trends, but we obviously were not successful in doing so.  We now lay out a more explicit approach and describe how we use magnitude of the slope, relative frequency of high/low years in the beginning and and end of the time frame, and the p-value together to assess for trends.  This is described in the paper in a new section in the methods, *Assessing regressions for trends*.   One of the biggest changes is to suggest values for the slope that would indicate an ecologically meaningful trend.  We describe this in the new section, but in general we are using a trophic state change from oligotrophic to eutrophic over one century to indicate ecologically meaningful change.  For temperature we look at values associated with ecological change as well as policy values (e.g. the Paris Agreement).  While these exact numbers are not universal, we use them as a starting point for assessing the trends.  How we interpreted the trends relative to the additional specific issues raised by the reviewer are detailed below and have changed somewhat from our initial submission, although the broad patterns still hold.

@@

- **Response:**  The results for TN in Rhode Island have changed somewhat since we added the additional filtering steps suggested by Reviewer 1 (**1-1** and **1-2**) and made changes to use the median instead of the mean in our response to comment **1-6**.  These changes have weakened the positive trend as the slope is smaller and the p-value larger.  Additionally three years, 1995, 1996, and 1997 now have insufficient data to be included in the figures and a third, 1998, has only three sites.  Using the more explicit definition of a trend we now conclude that TN showed neither a statistically positive trend or an ecologically important trend over the time frame of a century as the slope is less than the amount needed to cause an trophic state change.  This is reflected in the manuscript.

@@

- **Response:** We applied the same within and across year filters to the LAGOSNE, used the median instead of the mean (see **1-6**) and these results have changed as well.  Given that and our method for interpreting the trends leads to the following:  LAGOSNE is showing almost no statistical support for a change in nitrogen with above and below average years showing little pattern.  The slope is much less than the 3.5 we identified.  Thus we conclude no trend in TN at the scale of the Region.  This is indicated in the manuscript.

@@

- **Response:** Again, with the changes to the filtering and the switch to the median the trend lines have slightly changed. TP is showing no trend with no statistical support, a very small slope and above/below average years are relatively evenly distributed.  We have indicated as such in the text.

@@

- **Response:** We were a little unclear about this comment.  Is it referring to coherence over time within a single variable or coherence between the variables?  We agree that for individual variables some years do show greater variability than others.  These increases in the yearly standard deviations are largely the result of a few unusual measurements in those years for those variables.  Essentially these are outliers, but ones that have no obvious reason to remove from the analysis so they are kept in.  Further, there is no pattern to these outliers as they are from different ponds, aren't seen across variables in the same years, and in the case of the LAGOSNE data, they come from different states.  There is only one exception to this and that is the high variability in 2011 for both TN and TP in Rhode Island.  That is driven by one sampling event where TN and TP were high for both.  Given the lack of a pattern, there is little additional inference we can draw from this, other than the analysis might be susceptible to outliers.  Given that our original analysis is based on the mean and std. dev. and that those statistics are sensitive to outliers we have decided, thanks to this comment, to change to median and the inter-quartile range, both of which are considerably more robust statistics in the presence of outliers.  With this change the swings in variance are no longer seen as would be expected with statistics that are robust to outliers.  Good catch, by the reviewer.

@@

- **Response:** We agree with the reviewer that lake scale trends and the drivers behind those trends are both very interesting questions and something our approach could contribute to; however, we feel that it is beyond the scope of this particular paper.  Given the changes we have made to the title and the stated goals of the paper we feel that adding this type of analysis would greatly expand the paper and take away from our primary purpose of looking at long term trends in Rhode Island and the Northeast/Midwest.  We do feel this is a very good suggestion for next steps and my co-authors and I are in active discussion on how to move forward with such an analysis.   In the Discussion at the the end of *Trends* we briefly mention this application and that the anomaly approach could also be used at the individual lake level to assess for trends.

### 2020-09-25 manuscript/ecosphere_resubmit/manuscript.Rmd
push before resubmit to ecosphere

```{r analysis, include=FALSE}
source(here::here("R/data_available.R"), encoding = "UTF-8") 
source(here::here("R/summary_stats.R"), encoding = "UTF-8")  
source(here::here("R/figures_anomaly_only.R"), encoding = "UTF-8")
```

@@

There are many different methods for analyzing time series data for trends.  Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling locations is how to identify a trend while there is variation within a sampling site as well as variation introduced by differing start years for sampling among the many sites. For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more eutrophic water bodies.  Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].  The general approach we used calculates site-specific deviations from a long-term median over a pre-determined reference period.  This is slightly different than the typical use with temperature anomalies as those usually use the mean instead of the median.  Many of the variable we looked at are non-normal and often have outliers (e.g. algal blooms).  In this case the median is preferred. Using the median with site-specific deviations allows all sites to be shifted to a common baseline and the deviations, or anomalies, indicate change over the specified reference period.  We refer to this method as "site-specific anomalies".

*Summarizing site-specific anomalies:* Methods for calculating the site-specific anomalies and the yearly medians are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly median anomalies.", display = "cite")`.  Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from <https://github.com/usepa/ri_wq_trends> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].

@@

After filtering and summarizing the data some years may not have sufficient number of sites to be included.  We chose to include years in the analysis if they had at least three sites, but years with small numbers of sites are rare and only occurred with the nutrient data very early in the time frame of our analysis for the URIWW data(`r tab_num("year_n_ww", display = "cite", caption = "Number of URIWW sites available for trend analysis by parameter and year.")`) and late in the time frame for the LAGOSNE data (`r tab_num("year_n_lagosne", display = "cite", caption = "Number of LAGOSNE sites available for trend analysis by parameter and year.")`).

@@

```{r np_thresh, include=FALSE, message=FALSE}
d_and_m_1992 <- read_csv("data/downing_mccauley_1992_table_3.csv")
np_thresh <- d_and_m_1992 %>%
  mutate(molar_np = (total_n/total_p) * 2.21) %>%
  group_by(n_lim) %>%
  summarize(median_np = round(median(molar_np),0))
```

@@

In general, there was little evidence to suggest broad regional changes in chlorophyll *a* as it showed a very weak positive trend, slight statistical support and above average years spread evenly throughout the time period (slope = `r signif(lagos_chla_rl$slope,2)`, p = `r signif(lagos_chla_rl$p.value, 2)`, `r fig_num(name = "chla_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Total nitrogen showed a slight decreasing trend with weak statistical support and less than the 3.5 μg/L that would result in a trophic state change over a century.  The last two years for which LAGOSNE had data on total nitrogen were from a relatively small number of sites (`r tab_num("year_n_lagosne", display = "cite")`). There is little support for a meaningful trend in TN at the regional scale (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`). Total phosphorus showed a very small increasing trend and there is statistical support for that trend.  The slope however suggests little support for an ecological meaningful change in total phosphorus (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`). Lastly, the TN:TP ratio showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`) as none of our criteria for a trend were met.  Although, there appears to be a non-linear trend that suggests increasing TN:TP early in the time period and decreasing TN:TP later, we feel we cannot effectively evaluate this though as total nitrogen measurements were rare in 2011 (`r tab_num("year_n_lagosne", display = "cite")`) and sites that had both nitrogen and phosphorus were even rarer.  This low median anomaly should be interpreted with this caveat in mind.   Taken together we feel our results largely match the findings of Oliver et al. [-@Oliver_2017] that there is stasis in lake nutrients and chlorophyll within the LAGOSNE region. Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from <https://github.com/usepa/ri_wq_trends> and are archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].

@@

As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs (`r tab_num(name = "trend_summ", display = "cite")`).  However, 2006 and 2009 stand out as not following this trend.  Average May-October air temperature for 2009 was third lowest on record for the 1993 to 2016 time period and also showed the lowest maximum temperatures [@noaatrends].  On the other hand, 2006 was not unusually cool (11th lowest out of 24 years) but it was the wettest year on record for this period [@noaatrends].  These unusual weather patterns may help explain why 2006 and 2009 did not follow the trend.

@@

There are several broader implications from the results of our analysis and of examining long-term water quality trends in general. First, as more long-term datasets become available, it is important for managers, stakeholders, and researchers to work together to better understand long-term water quality trends at multiple spatial scales.  Specifically for this study, the results provide feedback to long time volunteer monitors about the trends in long-term, volunteer collected data.  This type of feedback is important in maintaining involvement as it has been shown in other areas that contributing to science and management is one of the expectations for some citizen science volunteers and can help maintain longer-term involvement [@ganzevoort2017sharing; @bonney2009citizen; @gouveia2004promoting].  Were that long-term involvement not common, then consistent long-term data may be more difficult to obtain and understanding of long-term trends would be made more difficult.

@@

Broad scale warming would be difficult to manage on a sub-regional scale thus other mechanisms, such as greater reductions in nutrients might compensate for warmer water temperatures and possible longer growing seasons.  Management efforts could focus on many approaches to further reduce nutrient loads such as source controls, enhanced entrainment of surface waters, treatment through green infrastructure, or in-lake approaches such as restoration of freshwater mussels [@yang2018nutrients; @kellogg2010geospatial; @pennino2016watershed; @reisinger2019changes; @kreeger2018restoration].  This is one possible hypothesis that is possible to form through information obtained by long-term trends. Additional work is certainly needed to explicitly examine the drivers of these trends in Rhode Island lakes and reservoirs.

>>>>

```{r analysis, include=FALSE, cache=TRUE}
source(here::here("R/data_clean.R"), encoding = "UTF-8")
source(here::here("R/data_available.R"), encoding = "UTF-8") 
source(here::here("R/summary_stats.R"), encoding = "UTF-8")  
source(here::here("R/figures_anomaly_only.R"), encoding = "UTF-8")
```

@@

There are many different methods for analyzing time series data for trends.  Environmental data are notoriously "noisy" and one of the difficulties that is encountered with multiple sampling locations is how to identify a trend while there is variation within a sampling site as well as variation introduced by differing start years for sampling among the many sites. For instance, if long-term data on water quality were collected more frequently in early years from more pristine waterbodies, then a simple comparison of raw values over time might show a decrease in water quality, which could be misleading if later sampling occurred on both pristine and more degreaded water bodies.  Thus, it is necessary to account for this type of within-site and among-site variation, using methods similar to those used to analyze long-term temperature trends using temperature anomalies [e.g., @jones1996calculating].  The general approach we used calculates site-specific deviations from a long-term median over a pre-determined reference period.  This is slightly different than the typical use with temperature anomalies as those usually use the mean instead of the median.  Many of the variable we looked at are non-normal and often have outliers (e.g. algal blooms).  In this case the median is preferred. Using the median with site-specific deviations allows all sites to be shifted to a common baseline and the deviations, or anomalies, indicate change over the specified reference period.  We refer to this method as "site-specific anomalies".

*Summarizing site-specific anomalies:* Methods for calculating the site-specific anomalies and the yearly medians are as follows and are presented graphically in `r fig_num(name = "anomaly_schematic", caption = "Example calculation of the site-specific anomalies and yearly median anomalies.", display = "cite")`.  Additionally, an example R script, `schematic_anomaly.R` and example dataset, `schematic.csv` to recreate and demonstrate the calculations in `r fig_num("anomaly_schematic", display = "cite")` is available from <https://github.com/usepa/ri_wq_trends/blob/master/R/schematic_anomaly.R> and is archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].

@@

After filtering and summarizing the data some years may not have sufficient number of sites to be included.  We chose to include years in the analysis if they had at least three sites, but years with small numbers of sites are rare and only occurred with the nutrient data very early in the time frame of our analysis for the URIWW data (`r tab_num("year_n_ww", display = "cite", caption = "Number of URIWW sites available for trend analysis by parameter and year.")`) and late in the time frame for the LAGOSNE data (`r tab_num("year_n_lagosne", display = "cite", caption = "Number of LAGOSNE sites available for trend analysis by parameter and year.")`).

@@

```{r np_thresh, include=FALSE, message=FALSE}
d_and_m_1992 <- read_csv(here::here("data/downing_mccauley_1992_table_3.csv"))
np_thresh <- d_and_m_1992 %>%
  mutate(molar_np = (total_n/total_p) * 2.21) %>%
  group_by(n_lim) %>%
  summarize(median_np = round(median(molar_np),0))
```

@@

In general, there was little evidence to suggest broad regional changes in chlorophyll *a* as it showed a very weak positive trend, slight statistical support and above average years spread evenly throughout the time period (slope = `r signif(lagos_chla_rl$slope,2)`, p = `r signif(lagos_chla_rl$p.value, 2)`, `r fig_num(name = "chla_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Total nitrogen showed a slight decreasing trend with weak statistical support and slope less than the 3.5 μg/L that would result in a trophic state change over a century, thus there is little support for a meaningful trend in TN at the regional scale (slope = `r signif(lagos_tn_rl$slope, 2)`, p = `r signif(lagos_tn_rl$p.value, 2)`, `r fig_num(name = "tn_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`).  Furthermore, The last two years for which LAGOSNE had data on total nitrogen were from a relatively small number of sites (`r tab_num("year_n_lagosne", display = "cite")`) and those values may not be representative.  Total phosphorus showed a very small increasing trend and there is statistical support for that trend.  The slope however suggests little support for an ecological meaningful change in total phosphorus (slope = `r signif(lagos_tp_rl$slope, 2)`, p = `r signif(lagos_tp_rl$p.value,2)`, `r fig_num(name = "tp_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`). Lastly, the TN:TP ratio showed little change (slope = `r signif(lagos_np_rl$slope, 2)`, p = `r signif(lagos_np_rl$p.value, 2)`, `r fig_num(name = "np_trend", display = "cite")`B., `r tab_num(name = "trend_summ", display = "cite")`) as none of our criteria for a trend were met.  Although, there appears to be a non-linear trend that suggests increasing TN:TP early in the time period and decreasing TN:TP later, we feel we cannot effectively evaluate this though as total nitrogen measurements were rare in 2011 (`r tab_num("year_n_lagosne", display = "cite")`) and sites that had both nitrogen and phosphorus were even rarer.  This low median anomaly should be interpreted with this caveat in mind.   Taken together we feel our results largely match the findings of Oliver et al. [-@Oliver_2017] that there is stasis in lake nutrients and chlorophyll within the LAGOSNE region. Data for all figures are available as a comma-separated values file, `yearly_average_anomaly.csv` from <https://github.com/USEPA/ri_wq_trends/blob/master/data/yearly_average_anomaly.csv> and are archived at <https://doi.org/10.5281/zenodo.3662828> [@zenodo_citation].

@@

As previously mentioned, both temperature and chlorophyll *a* show increasing trends from 1993 to 2016 in Rhode Island lakes and reservoirs (`r tab_num(name = "trend_summ", display = "cite")`).  However, 2006 and 2009 stand out as not following this trend for temperature.  Average May-October air temperature for 2009 was third lowest on record for the 1993 to 2016 time period and also showed the lowest maximum temperatures [@noaatrends].  On the other hand, 2006 was not unusually cool (11th lowest out of 24 years) but it was the wettest year on record for this period [@noaatrends].  These unusual weather patterns may help explain why 2006 and 2009 did not follow the trend.

@@

There are several broader implications from the results of our analysis and of examining long-term water quality trends in general. First, as more long-term datasets become available, it is important for managers, stakeholders, and researchers to work together to better understand long-term water quality trends at multiple spatial scales.  Specifically for this study, the results provide feedback to long time volunteer monitors about the trends in long-term, volunteer collected data.  This type of feedback is important in maintaining involvement as it has been shown in other areas that contributing to science and management is one of the expectations for some citizen science volunteers and can help maintain longer-term involvement [@ganzevoort2017sharing; @bonney2009citizen; @gouveia2004promoting].  If long-term involvement was not common, then consistent long-term data may be more difficult to obtain and understanding of long-term trends would be made more difficult.

@@

One possible hypothesis is that broad scale warming is difficult to manage on a sub-regional scale thus other mechanisms and approaches, such as greater reductions in nutrients might compensate for warmer water temperatures and possible longer growing seasons.  Management efforts could focus on many approaches to further reduce nutrient loads such as source controls, enhanced entrainment of surface waters, treatment through green infrastructure, or in-lake approaches such as restoration of freshwater mussels [@yang2018nutrients; @kellogg2010geospatial; @pennino2016watershed; @reisinger2019changes; @kreeger2018restoration].  There are of course many other hypotheses that we could form through information obtained by long-term trends such as those reported here. Additional work is certainly needed to explicitly examine the drivers of these trends in Rhode Island lakes and reservoirs and turn that information into management action.
